---
title: "Clustering-Based Sector Rotation with Kalman Filter Signals"
subtitle: "FIN 452 — Quantitative Trading Model"
author: "Nursultan Amangeldiev"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    fig-dpi: 150
    theme: cosmo
execute:
  cache: true
  message: false
  warning: false
  echo: true
---

```{r setup}
#| cache: false

# ── Packages ──────────────────────────────────────────────────────────────────
library(tidyverse)
library(tidyquant)
library(TTR)
library(tidymodels)
library(tidyclust)
library(factoextra)
library(KFAS)
library(PerformanceAnalytics)
library(plotly)
library(patchwork)
library(gt)
library(lubridate)
library(scales)
library(cluster)
library(rlang)
library(timetk)

# ── Global Constants ──────────────────────────────────────────────────────────
TICKERS <- c("SPY", "XLB", "XLC", "XLE", "XLF",
             "XLI", "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")

SECTOR_NAMES <- c(
  SPY  = "S&P 500 (Benchmark)",
  XLB  = "Materials",
  XLC  = "Communication Services",
  XLE  = "Energy",
  XLF  = "Financials",
  XLI  = "Industrials",
  XLK  = "Technology",
  XLP  = "Consumer Staples",
  XLRE = "Real Estate",
  XLV  = "Health Care",
  XLU  = "Utilities",
  XLY  = "Consumer Discretionary"
)

TRAIN_START <- as.Date("2000-06-01")
TRAIN_END   <- as.Date("2024-12-31")
TEST_START  <- as.Date("2025-01-01")
TEST_END    <- as.Date("2026-03-31")

N_CLUSTERS       <- 3L       # justified via elbow plot in Section 4
MOMENTUM_WINDOW  <- 6L       # months for favored-cluster labeling
KF_INNOV_CLIP_SD <- 3.0      # clip KF innovations beyond ±3 SD
TC_BPS           <- 0.0010   # 10 bps one-way transaction cost

set.seed(452)
```

```{r pkg-versions}
#| cache: false
#| tbl-cap: "Package versions used in this analysis"

pkgs <- c("tidyverse", "tidyquant", "TTR", "tidymodels", "tidyclust",
          "factoextra", "KFAS", "PerformanceAnalytics", "plotly",
          "patchwork", "gt", "lubridate", "scales", "timetk")

tibble::tibble(
  Package = pkgs,
  Version = purrr::map_chr(pkgs, \(p) as.character(packageVersion(p)))
) |>
  gt::gt() |>
  gt::tab_header(title = "R Package Versions") |>
  gt::cols_align("left")
```

# Introduction & Strategy Overview

This report implements a **clustering-based sector rotation strategy** applied to the eleven S&P 500 Select Sector SPDR ETFs, using SPY as a benchmark. The strategy is a long-biased active fund with a short overlay, targeting returns above the index by going long on the strongest sectors and shorting the weakest. It combines two complementary methods:

1. **K-Means Clustering** groups sectors by their return and momentum characteristics each month. The cluster with the strongest trailing momentum is the *favored* cluster (long-eligible); the cluster with the weakest is the *unfavored* cluster (short-eligible).

2. **Kalman Filter** fits a local-level state-space model to each sector's return history. The one-step-ahead *innovation* (observed minus predicted return) measures whether a sector is outperforming or underperforming its recent trend. The sign and magnitude of the innovation triggers and sizes the trade.

**Signal combination rule:**

| Condition | Position |
|-----------|----------|
| Favored cluster + positive KF innovation | **Long** (weights sum to +100%) |
| Unfavored cluster + negative KF innovation | **Short** (weights sum to −30%) |
| All other sectors | **Flat** (0%) |

Net portfolio exposure is **+70%**, consistent with a long-biased active fund. Monthly rebalancing occurs on the last trading day of each month, with execution assumed at the same day's closing price. A 10 bps one-way transaction cost is applied proportional to weight changes.

**Training period:** June 2000 – December 2024
**Testing period:** January 2025 – March 2026

# Data Acquisition & Validation

## Pulling Price Data

All data is sourced from Yahoo Finance via the `tidyquant` package. No local data files are used. Adjusted closing prices account for dividends and splits.

```{r pull-prices}
prices_raw <- tidyquant::tq_get(
  TICKERS,
  from = TRAIN_START,
  to   = TEST_END,
  get  = "stock.prices"
)
```

## Ticker Availability

XLC (Communication Services) launched on 2018-06-19 and XLRE (Real Estate) on 2015-10-07. These tickers return fewer observations than the others. An eligibility check ensures they are excluded from portfolio allocation before their respective inception dates.

```{r ticker-availability}
#| tbl-cap: "Observation counts and first available date per ticker"

ticker_meta <- prices_raw |>
  dplyr::group_by(symbol) |>
  dplyr::summarise(
    first_date = min(date),
    last_date  = max(date),
    n_obs      = dplyr::n(),
    .groups    = "drop"
  ) |>
  dplyr::mutate(sector = SECTOR_NAMES[symbol])

ticker_meta |>
  dplyr::select(symbol, sector, first_date, last_date, n_obs) |>
  gt::gt() |>
  gt::tab_header(title = "Ticker Availability") |>
  gt::cols_label(
    symbol     = "Ticker",
    sector     = "Sector",
    first_date = "First Date",
    last_date  = "Last Date",
    n_obs      = "Observations"
  ) |>
  gt::cols_align("left")
```

## Data Quality Checks

We screen for: (1) NA adjusted prices, (2) zero-volume days, and (3) single-day price moves exceeding 50% (likely data errors despite using adjusted prices).

```{r data-quality}
#| tbl-cap: "Data quality flags per ticker"

quality_flags <- prices_raw |>
  dplyr::group_by(symbol) |>
  dplyr::summarise(
    na_price      = sum(is.na(adjusted)),
    zero_volume   = sum(volume == 0, na.rm = TRUE),
    extreme_moves = sum(
      abs(adjusted / dplyr::lag(adjusted) - 1) > 0.50,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

quality_flags |>
  gt::gt() |>
  gt::tab_header(title = "Data Quality Flags") |>
  gt::cols_label(
    symbol        = "Ticker",
    na_price      = "NA Prices",
    zero_volume   = "Zero-Volume Days",
    extreme_moves = ">50% Daily Move"
  ) |>
  gt::data_color(
    columns = c(na_price, zero_volume, extreme_moves),
    method  = "numeric",
    palette = c("white", "orange")
  )
```

## Daily Returns

```{r daily-returns}
returns_daily <- prices_raw |>
  dplyr::group_by(symbol) |>
  tidyquant::tq_transmute(
    select     = adjusted,
    mutate_fun = periodReturn,
    period     = "daily",
    col_rename = "daily_ret"
  ) |>
  dplyr::ungroup()
```

## Monthly Returns

Monthly returns are used for rebalancing signal computation and portfolio performance tracking. We use end-of-month closing prices.

```{r monthly-returns}
returns_monthly <- prices_raw |>
  dplyr::group_by(symbol) |>
  tidyquant::tq_transmute(
    select     = adjusted,
    mutate_fun = periodReturn,
    period     = "monthly",
    col_rename = "monthly_ret"
  ) |>
  dplyr::ungroup() |>
  # date is set to last trading day of each month by tq_transmute
  dplyr::rename(month_end = date)
```

## Price History Visualization

```{r price-chart}
#| fig-cap: "Adjusted price history for all tickers (rebased to 100 at each series start)"

rebased <- prices_raw |>
  dplyr::group_by(symbol) |>
  dplyr::mutate(rebased = adjusted / first(adjusted) * 100) |>
  dplyr::ungroup() |>
  dplyr::mutate(sector = SECTOR_NAMES[symbol])

p <- ggplot2::ggplot(rebased, ggplot2::aes(x = date, y = rebased, color = symbol)) +
  ggplot2::geom_line(linewidth = 0.5, alpha = 0.85) +
  ggplot2::geom_vline(xintercept = as.numeric(TEST_START),
             linetype = "dashed", color = "grey40") +
  ggplot2::annotate("text", x = TEST_START, y = max(rebased$rebased, na.rm = TRUE) * 0.95,
           label = "Test period →", hjust = -0.1, size = 3.5, color = "grey40") +
  ggplot2::scale_y_continuous(labels = scales::label_comma()) +
  ggplot2::scale_color_brewer(palette = "Paired") +
  ggplot2::labs(title = "Rebased Price History (base = 100 at inception)",
       x = NULL, y = "Rebased Price", color = "Ticker") +
  ggplot2::theme_minimal(base_size = 12)

plotly::ggplotly(p, tooltip = c("x", "y", "colour")) |>
  plotly::layout(hovermode = "x unified")
```

```{r returns-summary}
#| tbl-cap: "Annualized return and volatility over the full training period"

train_summary <- returns_daily |>
  dplyr::filter(date >= TRAIN_START, date <= TRAIN_END) |>
  dplyr::group_by(symbol) |>
  dplyr::summarise(
    ann_return = prod(1 + daily_ret, na.rm = TRUE)^(252 / dplyr::n()) - 1,
    ann_vol    = stats::sd(daily_ret, na.rm = TRUE) * sqrt(252),
    sharpe     = ann_return / ann_vol,
    .groups    = "drop"
  ) |>
  dplyr::mutate(sector = SECTOR_NAMES[symbol]) |>
  dplyr::arrange(dplyr::desc(ann_return))

train_summary |>
  dplyr::select(symbol, sector, ann_return, ann_vol, sharpe) |>
  gt::gt() |>
  gt::tab_header(title = "Training Period Summary (Jun 2000 – Dec 2024)") |>
  gt::cols_label(
    symbol     = "Ticker",
    sector     = "Sector",
    ann_return = "Ann. Return",
    ann_vol    = "Ann. Volatility",
    sharpe     = "Sharpe (rf=0)"
  ) |>
  gt::fmt_percent(columns = c(ann_return, ann_vol), decimals = 1) |>
  gt::fmt_number(columns = sharpe, decimals = 2) |>
  gt::data_color(
    columns = sharpe,
    method  = "numeric",
    palette = c("tomato", "white", "steelblue")
  )
```

# Feature Engineering

Features are computed on daily price data and sampled at each month-end. The clustering unit is a **ticker × month** observation: each row represents one sector at one point in time, described by a vector of indicators. All normalization parameters are estimated on the training set only and applied to the test set — preventing look-ahead bias.

## Computing Indicators

We compute seven features per ticker per day using `TTR`:

| Feature | Description |
|---------|-------------|
| `mom_1m` | 1-month trailing return |
| `mom_3m` | 3-month trailing return |
| `mom_6m` | 6-month trailing return |
| `mom_12m` | 12-month trailing return |
| `trend_ratio` | 50-day SMA ÷ 200-day SMA (price relative to long-term trend) |
| `vol_20d` | 20-day realized annualized volatility |
| `rsi_14` | 14-day Relative Strength Index |

Momentum features capture cross-sectional relative strength. The trend ratio flags whether a sector is in a short-term uptrend relative to its own history. Volatility and RSI add regime context that clustering can exploit.

```{r compute-indicators}
# Helper: compute all features for a single ticker's price series
compute_features <- function(df) {
  df <- df |> dplyr::arrange(date)
  n  <- nrow(df)
  p  <- df$adjusted

  # Rolling returns at various horizons (approximate trading days)
  ret_1m  <- p / dplyr::lag(p, 21)  - 1
  ret_3m  <- p / dplyr::lag(p, 63)  - 1
  ret_6m  <- p / dplyr::lag(p, 126) - 1
  ret_12m <- p / dplyr::lag(p, 252) - 1

  sma50  <- TTR::SMA(p, n = 50)
  sma200 <- TTR::SMA(p, n = 200)

  # Daily log returns for realized vol
  log_ret <- c(NA, diff(log(p)))
  vol_20d <- zoo::rollapply(log_ret, width = 20, FUN = stats::sd, fill = NA, align = "right") * sqrt(252)

  rsi_14 <- TTR::RSI(p, n = 14)

  df |>
    dplyr::mutate(
      mom_1m      = ret_1m,
      mom_3m      = ret_3m,
      mom_6m      = ret_6m,
      mom_12m     = ret_12m,
      trend_ratio = sma50 / sma200,
      vol_20d     = vol_20d,
      rsi_14      = rsi_14
    ) |>
    # symbol is stripped by group_modify — it is re-attached automatically
    dplyr::select(date, mom_1m, mom_3m, mom_6m, mom_12m,
                  trend_ratio, vol_20d, rsi_14)
}

features_daily <- prices_raw |>
  dplyr::group_by(symbol) |>
  dplyr::group_modify(~ compute_features(.x)) |>
  dplyr::ungroup()
```

## Sampling at Month-End

We join daily features to the monthly return dates to get one feature snapshot per ticker per month-end rebalancing date.

```{r month-end-features}
# Month-end dates from the returns_monthly table (covers full date range)
month_ends <- returns_monthly |>
  dplyr::select(symbol, month_end) |>
  dplyr::distinct()

features_monthly <- features_daily |>
  dplyr::inner_join(month_ends, by = c("symbol", "date" = "month_end")) |>
  dplyr::rename(month_end = date) |>
  dplyr::arrange(symbol, month_end)
```

## Train / Test Split

```{r feature-split}
features_train <- features_monthly |>
  dplyr::filter(month_end >= TRAIN_START, month_end <= TRAIN_END)

features_test <- features_monthly |>
  dplyr::filter(month_end >= TEST_START, month_end <= TEST_END)

cat(
  "Training observations:", nrow(features_train),
  "\nTest observations:    ", nrow(features_test)
)
```

## Tidymodels Recipe

The recipe imputes missing values first (median of each feature, computed on training data), then normalizes all features to zero mean and unit variance. This order matters: imputing before normalization means the imputed value equals the training median, which after scaling maps to 0 — the most neutral possible value for a missing observation.

```{r feature-recipe}
FEATURE_COLS <- c("mom_1m", "mom_3m", "mom_6m", "mom_12m",
                  "trend_ratio", "vol_20d", "rsi_14")

cluster_recipe <- recipes::recipe(~ ., data = features_train) |>
  recipes::update_role(symbol, month_end, new_role = "id") |>
  recipes::step_impute_median(all_of(FEATURE_COLS)) |>
  recipes::step_normalize(all_of(FEATURE_COLS))

# Prep on training data only
cluster_recipe_prepped <- recipes::prep(cluster_recipe, training = features_train)

# Bake both sets using training statistics
features_train_baked <- recipes::bake(cluster_recipe_prepped, new_data = features_train)
features_test_baked  <- recipes::bake(cluster_recipe_prepped, new_data = features_test)
```

## Feature Distributions

```{r feature-distributions}
#| fig-cap: "Distribution of normalized features across all sectors and training months"
#| fig-height: 7

features_train_baked |>
  tidyr::pivot_longer(all_of(FEATURE_COLS), names_to = "feature", values_to = "value") |>
  dplyr::mutate(feature = factor(feature, levels = FEATURE_COLS)) |>
  ggplot2::ggplot(ggplot2::aes(x = value, fill = feature)) +
  ggplot2::geom_histogram(bins = 40, color = "white", linewidth = 0.2, show.legend = FALSE) +
  ggplot2::facet_wrap(~ feature, scales = "free_y", ncol = 2) +
  ggplot2::labs(title = "Normalized Feature Distributions (Training Set)",
       x = "Normalized Value", y = "Count") +
  ggplot2::theme_minimal(base_size = 11)
```

## Feature Correlation

```{r feature-correlation}
#| fig-cap: "Pairwise correlation of normalized features (training set)"
#| fig-height: 6

cor_mat <- features_train_baked |>
  dplyr::select(all_of(FEATURE_COLS)) |>
  stats::cor(use = "pairwise.complete.obs")

cor_df <- cor_mat |>
  as.data.frame() |>
  tibble::rownames_to_column("feature_x") |>
  tidyr::pivot_longer(-feature_x, names_to = "feature_y", values_to = "correlation") |>
  dplyr::mutate(
    feature_x = factor(feature_x, levels = FEATURE_COLS),
    feature_y = factor(feature_y, levels = rev(FEATURE_COLS))
  )

ggplot2::ggplot(cor_df, ggplot2::aes(x = feature_x, y = feature_y, fill = correlation)) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::geom_text(ggplot2::aes(label = round(correlation, 2)), size = 3.5) +
  ggplot2::scale_fill_gradient2(low = "tomato", mid = "white", high = "steelblue",
                       midpoint = 0, limits = c(-1, 1)) +
  ggplot2::labs(title = "Feature Correlation Matrix (Training Set)",
       x = NULL, y = NULL, fill = "Correlation") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))
```

# Clustering Model

We cluster the 11 sector ETFs each month using K-Means on the normalized feature vector. Clustering tickers (rather than time periods) produces a per-sector regime label at each rebalancing date — the natural input for a sector rotation strategy.

**Key design choices:**

- K-Means is re-fit each month on the **full training history up to that month** (expanding window), so cluster assignments are always out-of-sample relative to the future.
- The "favored" cluster is labeled dynamically each month as the cluster with the **highest average 6-month trailing momentum** (`mom_6m`) among its members — not by a hardcoded cluster number, since K-Means label numbering is arbitrary.
- The "unfavored" cluster is the one with the **lowest average `mom_6m`**. The middle cluster (if K=3) is treated as neutral (flat).

## Hypothesis: Expected Cluster Structure

Before examining the data, financial theory gives us a prior on how sectors should group:

| Expected Cluster | Sectors | Rationale |
|---|---|---|
| **Defensive / Low-volatility** | XLU, XLP, XLV | Low beta, stable cash flows, bond-like; outperform in late-cycle and recessions |
| **Cyclical / Growth** | XLK, XLI, XLY, XLB | High beta, earnings sensitive to economic expansion; outperform early-cycle |
| **Interest-rate sensitive** | XLF, XLRE, XLC | Driven by credit conditions and rate environment more than pure economic cycle |
| **Commodity / Energy** | XLE | Commodity price exposure creates a unique return driver distinct from other sectors |

> Our approach: let the data reveal the structure, then validate against economic intuition. Agreement between data and theory increases confidence in the cluster structure.

## Selecting K: Elbow & Silhouette Analysis

We evaluate K = 2 through 6 on the full training baked features. Both the within-cluster sum of squares (elbow) and the average silhouette width inform — but do not mechanically determine — the choice of K.

```{r select-k}
#| fig-cap: "Elbow and silhouette plots for K-Means cluster count selection"
#| fig-height: 5

# Use the full training baked matrix for K selection
X_train <- features_train_baked |>
  dplyr::select(all_of(FEATURE_COLS)) |>
  tidyr::drop_na()

set.seed(452)
k_range <- 2:6

# Within-cluster SS and silhouette per K
k_metrics <- purrr::map_dfr(k_range, function(k) {
  km  <- stats::kmeans(X_train, centers = k, nstart = 25, iter.max = 100)
  sil <- cluster::silhouette(km$cluster, stats::dist(X_train))
  tibble::tibble(
    k         = k,
    tot_withinss = km$tot.withinss,
    avg_sil   = mean(sil[, 3])
  )
})

p_elbow <- ggplot2::ggplot(k_metrics, ggplot2::aes(x = k, y = tot_withinss)) +
  ggplot2::geom_line(color = "steelblue", linewidth = 1) +
  ggplot2::geom_point(size = 3, color = "steelblue") +
  ggplot2::geom_vline(xintercept = N_CLUSTERS, linetype = "dashed", color = "tomato") +
  ggplot2::labs(title = "Elbow Plot", x = "Number of Clusters (K)",
       y = "Total Within-Cluster SS") +
  ggplot2::theme_minimal(base_size = 11)

p_sil <- ggplot2::ggplot(k_metrics, ggplot2::aes(x = k, y = avg_sil)) +
  ggplot2::geom_line(color = "darkgreen", linewidth = 1) +
  ggplot2::geom_point(size = 3, color = "darkgreen") +
  ggplot2::geom_vline(xintercept = N_CLUSTERS, linetype = "dashed", color = "tomato") +
  ggplot2::labs(title = "Average Silhouette Width", x = "Number of Clusters (K)",
       y = "Avg Silhouette") +
  ggplot2::theme_minimal(base_size = 11)

p_elbow + p_sil +
  patchwork::plot_annotation(
    title   = "K Selection: Elbow and Silhouette (red dashed = chosen K)",
    theme   = ggplot2::theme_minimal(base_size = 12)
  )
```

```{r k-metrics-table}
#| tbl-cap: "Within-cluster SS and silhouette width by K"

k_metrics |>
  gt::gt() |>
  gt::cols_label(k = "K", tot_withinss = "Total Within-SS", avg_sil = "Avg Silhouette") |>
  gt::fmt_number(columns = c(tot_withinss, avg_sil), decimals = 3) |>
  gt::tab_style(
    style = list(gt::cell_fill(color = "#d4edda")),
    locations = gt::cells_body(rows = k == N_CLUSTERS)
  )
```

> **Choice: K = `r N_CLUSTERS`.** The silhouette plot peaks at K = 2 (≈ 0.37), which is the statistically cleanest partition. However, K = 2 is rejected on **strategic grounds**: with only two clusters, every sector must be either Long or Short — there is no neutral buffer. The Long/Flat/Short signal architecture explicitly requires a third state (Flat), which needs a distinct "neutral" cluster to hold sectors that are neither the strongest nor the weakest at a given month-end. Eliminating that buffer would force the strategy into a fully invested long-short position at all times, which contradicts the design intent and increases unnecessary turnover during ambiguous regimes. K = 3 is therefore the minimum K consistent with the three-state signal. The elbow shows a meaningful compactness gain moving from K = 2 to K = 3 (~19% reduction in within-cluster SS), after which gains diminish — consistent with K = 3 as a reasonable stopping point. K ≥ 4 is also rejected: with only 11 tickers per monthly cross-section, K = 4 or higher risks singleton or near-singleton clusters, which are statistically unreliable and produce degenerate portfolio weights. The three-cluster structure also maps directly onto well-established economic regime archetypes: **Growth** (high momentum, high trend-ratio), **Defensive** (low volatility, stable RSI), and **Cyclical** (high volatility, mixed momentum) — labels that are economically interpretable and defensible.

## Fitting the Cluster Model (tidymodels / tidyclust)

We fit a single K-Means model on the full training data for EDA and visualization. For portfolio signal generation (Section 7), we re-fit monthly on expanding windows to avoid look-ahead bias.

```{r fit-kmeans}
set.seed(452)

kmeans_spec <- tidyclust::k_means(num_clusters = N_CLUSTERS) |>
  parsnip::set_engine("stats")

cluster_wflow <- workflows::workflow() |>
  workflows::add_recipe(cluster_recipe) |>
  workflows::add_model(kmeans_spec)

cluster_fit <- parsnip::fit(cluster_wflow, data = features_train)

# Extract cluster assignments on training data
train_assignments <- features_train |>
  dplyr::bind_cols(
    cluster = tidyclust::extract_cluster_assignment(cluster_fit)$.cluster
  )
```

## Dynamic Cluster Labeling

Cluster numbers from K-Means are arbitrary. We label them by their average `mom_6m` centroid value each month: the cluster with the highest average 6-month momentum is **Favored** (long-eligible); the lowest is **Unfavored** (short-eligible); the middle is **Neutral**.

```{r cluster-labels}
# Compute centroid mom_6m per cluster across full training assignments
cluster_centroids <- train_assignments |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(
    mean_mom_6m     = mean(mom_6m, na.rm = TRUE),
    mean_mom_1m     = mean(mom_1m, na.rm = TRUE),
    mean_vol        = mean(vol_20d, na.rm = TRUE),
    mean_rsi        = mean(rsi_14, na.rm = TRUE),
    mean_trend      = mean(trend_ratio, na.rm = TRUE),
    n_obs           = dplyr::n(),
    .groups         = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(mean_mom_6m)) |>
  dplyr::mutate(label = c("Favored (Growth)", "Neutral", "Unfavored (Defensive/Laggard)")[
    rank(-mean_mom_6m, ties.method = "first")
  ])

cluster_centroids |>
  dplyr::select(cluster, label, mean_mom_6m, mean_mom_1m, mean_vol, mean_rsi, mean_trend, n_obs) |>
  gt::gt() |>
  gt::tab_header(title = "Cluster Centroids (Full Training Period)") |>
  gt::cols_label(
    cluster      = "Cluster",
    label        = "Label",
    mean_mom_6m  = "Avg 6M Mom",
    mean_mom_1m  = "Avg 1M Mom",
    mean_vol     = "Avg Vol (Ann.)",
    mean_rsi     = "Avg RSI",
    mean_trend   = "Avg Trend Ratio",
    n_obs        = "Obs"
  ) |>
  gt::fmt_percent(columns = c(mean_mom_6m, mean_mom_1m, mean_vol), decimals = 1) |>
  gt::fmt_number(columns = c(mean_rsi, mean_trend), decimals = 2) |>
  gt::tab_style(
    style = list(gt::cell_fill(color = "#cce5ff")),
    locations = gt::cells_body(rows = label == "Favored (Growth)")
  ) |>
  gt::tab_style(
    style = list(gt::cell_fill(color = "#f8d7da")),
    locations = gt::cells_body(rows = label == "Unfavored (Defensive/Laggard)")
  )
```

## PCA Visualization of Clusters

```{r pca-cluster-plot}
#| fig-cap: "PCA biplot of sector-month observations colored by cluster label"
#| fig-height: 6

# features_train_baked has no NAs (imputation was applied in the recipe)
# Join cluster assignments by (symbol, month_end) — avoid positional bind_cols
pca_input <- features_train_baked |>
  dplyr::left_join(
    train_assignments |> dplyr::select(symbol, month_end, cluster),
    by = c("symbol", "month_end")
  ) |>
  tidyr::drop_na(all_of(FEATURE_COLS))

pca_result <- stats::prcomp(
  pca_input |> dplyr::select(all_of(FEATURE_COLS)),
  center = FALSE, scale. = FALSE  # already normalized
)

# Map cluster numbers to labels using centroid table
label_map <- cluster_centroids |> dplyr::select(cluster, label) |> tibble::deframe()

pca_df <- tibble::as_tibble(pca_result$x[, 1:2]) |>
  dplyr::bind_cols(
    pca_input |> dplyr::select(symbol, month_end, cluster)
  ) |>
  dplyr::mutate(label = label_map[as.character(cluster)])

var_explained <- summary(pca_result)$importance[2, 1:2] * 100

ggplot2::ggplot(pca_df, ggplot2::aes(x = PC1, y = PC2, color = label, label = symbol)) +
  ggplot2::geom_point(alpha = 0.4, size = 1.8) +
  ggplot2::stat_ellipse(linewidth = 0.8, level = 0.75) +
  ggplot2::scale_color_manual(values = c(
    "Favored (Growth)"             = "steelblue",
    "Neutral"                      = "grey50",
    "Unfavored (Defensive/Laggard)" = "tomato"
  )) +
  ggplot2::labs(
    title = "K-Means Clusters in PCA Space (Training Period)",
    x     = sprintf("PC1 (%.1f%% variance)", var_explained[1]),
    y     = sprintf("PC2 (%.1f%% variance)", var_explained[2]),
    color = "Cluster"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

## Cluster Composition Over Time

Which sectors end up in which cluster, and how stable are assignments over time?

```{r cluster-heatmap}
#| fig-cap: "Cluster assignment per sector over time (training period)"
#| fig-height: 6

train_assignments |>
  dplyr::mutate(label = label_map[as.character(cluster)]) |>
  ggplot2::ggplot(ggplot2::aes(x = month_end, y = symbol, fill = label)) +
  ggplot2::geom_tile(color = "white", linewidth = 0.3) +
  ggplot2::scale_fill_manual(values = c(
    "Favored (Growth)"             = "steelblue",
    "Neutral"                      = "grey80",
    "Unfavored (Defensive/Laggard)" = "tomato"
  )) +
  ggplot2::labs(
    title = "Cluster Assignment by Sector and Month (Training Period)",
    x = NULL, y = NULL, fill = "Cluster"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## Hierarchical Clustering (Robustness Check)

K-Means assumes spherical, equally-sized clusters. Hierarchical clustering (Ward's method) makes no such assumption and produces a dendrogram that reveals the natural nesting structure of sectors. Agreement between both methods signals a robust, data-driven structure. Disagreement warrants further investigation.

```{r hclust-fit}
#| fig-cap: "Dendrogram of sector ETFs — hierarchical clustering with Ward's method (full training period)"
#| fig-height: 6

# Use the same normalized full-training feature matrix
X_train_labeled <- features_train_baked |>
  dplyr::select(all_of(FEATURE_COLS)) |>
  tidyr::drop_na()

# Aggregate to one representative vector per ticker (median across all months)
X_train_by_ticker <- features_train_baked |>
  tidyr::drop_na(all_of(FEATURE_COLS)) |>
  dplyr::group_by(symbol) |>
  dplyr::summarise(dplyr::across(all_of(FEATURE_COLS), stats::median), .groups = "drop")

ticker_labels <- X_train_by_ticker$symbol
X_hclust <- X_train_by_ticker |> dplyr::select(all_of(FEATURE_COLS)) |> as.matrix()
rownames(X_hclust) <- ticker_labels

dist_mat    <- stats::dist(X_hclust, method = "euclidean")
hclust_fit  <- stats::hclust(dist_mat, method = "ward.D2")

# Cut tree at K = N_CLUSTERS
hclust_labels <- stats::cutree(hclust_fit, k = N_CLUSTERS)

factoextra::fviz_dend(
  hclust_fit,
  k            = N_CLUSTERS,
  cex          = 0.9,
  horiz        = TRUE,
  rect         = TRUE,
  rect_fill    = TRUE,
  rect_border  = c("steelblue", "grey50", "tomato"),
  main         = "Hierarchical Clustering of Sector ETFs (Ward's Method)",
  sub          = sprintf("Training period %s – %s | k = %d", TRAIN_START, TRAIN_END, N_CLUSTERS)
)
```

> **Reading the dendrogram:** Height = distance at which clusters merge. Sectors that merge at low height are most similar. Rectangles show the K = `r N_CLUSTERS` cut. Sectors in the same box belong to the same cluster.

## K-Means vs. Hierarchical: Method Agreement

If both methods produce the same groupings, the cluster structure is robust. If they disagree on a sector, that sector sits near a cluster boundary.

```{r method-comparison}
#| tbl-cap: "Cluster assignments from K-Means and Hierarchical clustering"

# K-Means label per ticker (majority vote across training months)
kmeans_by_ticker <- train_assignments |>
  dplyr::group_by(symbol) |>
  dplyr::summarise(
    kmeans_cluster = as.integer(names(sort(table(cluster), decreasing = TRUE))[1]),
    .groups = "drop"
  ) |>
  dplyr::mutate(kmeans_label = label_map[as.character(kmeans_cluster)])

# Hierarchical label per ticker
hclust_by_ticker <- tibble::tibble(
  symbol         = names(hclust_labels),
  hclust_cluster = hclust_labels
)

# Join and compare
method_comparison <- kmeans_by_ticker |>
  dplyr::left_join(hclust_by_ticker, by = "symbol") |>
  dplyr::mutate(
    agree = kmeans_cluster == hclust_cluster,
    sector = SECTOR_NAMES[symbol]
  ) |>
  dplyr::arrange(kmeans_cluster)

method_comparison |>
  dplyr::select(symbol, sector, kmeans_label, hclust_cluster, agree) |>
  gt::gt() |>
  gt::tab_header(
    title    = "K-Means vs. Hierarchical Clustering",
    subtitle = "Agreement → robust structure | Disagreement → boundary sector"
  ) |>
  gt::cols_label(
    symbol         = "Ticker",
    sector         = "Sector",
    kmeans_label   = "K-Means Label",
    hclust_cluster = "Hierarchical Cluster",
    agree          = "Agree?"
  ) |>
  gt::tab_style(
    style     = list(gt::cell_fill(color = "#d4edda")),
    locations = gt::cells_body(rows = agree == TRUE)
  ) |>
  gt::tab_style(
    style     = list(gt::cell_fill(color = "#f8d7da")),
    locations = gt::cells_body(rows = agree == FALSE)
  )

n_agree <- sum(method_comparison$agree, na.rm = TRUE)
n_total <- nrow(method_comparison)
```

> **`r n_agree` of `r n_total` sectors agree** between K-Means and Hierarchical clustering. `r if(n_agree >= 9) "High agreement confirms a robust cluster structure." else if(n_agree >= 7) "Moderate agreement — boundary sectors should be interpreted cautiously." else "Low agreement — cluster boundaries are soft; interpret with caution."`

## Cluster Stability Across Regimes

Clusters that are stable across very different market regimes (pre-COVID expansion vs. post-COVID recovery) are more reliable for a live trading strategy. We split the training period at 2020-01-01 and compare cluster assignments before and after.

```{r stability-by-regime}
#| fig-cap: "Average feature profile by sector: Pre-COVID (2000–2019) vs. Post-COVID (2020–2024)"
#| fig-height: 5

regime_features <- features_train_baked |>
  dplyr::mutate(regime = dplyr::if_else(month_end < as.Date("2020-01-01"), "Pre-COVID (2000–2019)", "Post-COVID (2020–2024)")) |>
  dplyr::group_by(symbol, regime) |>
  dplyr::summarise(dplyr::across(all_of(FEATURE_COLS), \(x) mean(x, na.rm = TRUE)), .groups = "drop")

regime_features |>
  tidyr::pivot_longer(all_of(FEATURE_COLS), names_to = "feature", values_to = "value") |>
  dplyr::filter(feature %in% c("mom_6m", "vol_20d", "rsi_14")) |>
  ggplot2::ggplot(ggplot2::aes(x = symbol, y = value, fill = regime)) +
  ggplot2::geom_col(position = "dodge", width = 0.7) +
  ggplot2::facet_wrap(~ feature, scales = "free_y", ncol = 1) +
  ggplot2::scale_fill_manual(values = c("Pre-COVID (2000–2019)" = "steelblue",
                               "Post-COVID (2020–2024)" = "tomato")) +
  ggplot2::labs(title = "Feature Profiles Before and After COVID",
       x = NULL, y = "Normalized Value", fill = "Regime") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))
```

```{r stability-cluster-compare}
#| tbl-cap: "K-Means majority cluster label per sector: Pre-COVID vs. Post-COVID"

# Re-cluster on each regime's feature set
cluster_for_regime <- function(data_baked) {
  X <- data_baked |> dplyr::select(all_of(FEATURE_COLS)) |> tidyr::drop_na() |>
    dplyr::group_by(data_baked |> tidyr::drop_na(all_of(FEATURE_COLS)) |> dplyr::pull(symbol)) |>
    dplyr::summarise(dplyr::across(dplyr::everything(), stats::median), .groups = "drop")
  # Return median feature per ticker
  data_baked |>
    tidyr::drop_na(all_of(FEATURE_COLS)) |>
    dplyr::group_by(symbol) |>
    dplyr::summarise(dplyr::across(all_of(FEATURE_COLS), stats::median), .groups = "drop")
}

pre  <- features_train_baked |> dplyr::filter(month_end < as.Date("2020-01-01"))
post <- features_train_baked |> dplyr::filter(month_end >= as.Date("2020-01-01"))

assign_clusters <- function(df, k = N_CLUSTERS) {
  mat <- df |> tidyr::drop_na(all_of(FEATURE_COLS)) |>
    dplyr::group_by(symbol) |>
    dplyr::summarise(dplyr::across(all_of(FEATURE_COLS), stats::median), .groups = "drop")
  set.seed(452)
  km <- stats::kmeans(mat |> dplyr::select(all_of(FEATURE_COLS)), centers = k, nstart = 25)
  # Label by mom_6m centroid rank
  centroid_mom <- km$centers[, "mom_6m"]
  rank_map     <- rank(-centroid_mom, ties.method = "first")
  labels_vec   <- c("Favored", "Neutral", "Unfavored")[rank_map]
  mat |> dplyr::transmute(symbol, cluster_label = labels_vec[km$cluster])
}

pre_clusters  <- assign_clusters(pre)  |> dplyr::rename(pre_label  = cluster_label)
post_clusters <- assign_clusters(post) |> dplyr::rename(post_label = cluster_label)

pre_clusters |>
  dplyr::left_join(post_clusters, by = "symbol") |>
  dplyr::mutate(
    stable = pre_label == post_label,
    sector = SECTOR_NAMES[symbol]
  ) |>
  dplyr::select(symbol, sector, pre_label, post_label, stable) |>
  gt::gt() |>
  gt::tab_header(title    = "Cluster Stability: Pre-COVID vs. Post-COVID",
             subtitle = "Stable sectors remain in the same cluster across both regimes") |>
  gt::cols_label(symbol = "Ticker", sector = "Sector",
             pre_label = "Pre-COVID", post_label = "Post-COVID", stable = "Stable?") |>
  gt::tab_style(style = list(gt::cell_fill(color = "#d4edda")),
            locations = gt::cells_body(rows = stable == TRUE)) |>
  gt::tab_style(style = list(gt::cell_fill(color = "#fff3cd")),
            locations = gt::cells_body(rows = stable == FALSE))
```

## When Clustering Fails: Monthly Health Check

The class notes identify five red flags that indicate clustering is unreliable in a given month. We flag these in the training data. Months with unreliable clustering will be treated as **flat (no position)** in the portfolio construction step.

**Red flags (any one triggers a warning):**
1. Average silhouette width < 0.3
2. Largest cluster contains > 70% of tickers (near-degenerate solution)
3. A cluster has only 1 member

```{r clustering-health}
#| fig-cap: "Monthly average silhouette width over the training period (dashed = 0.3 threshold)"
#| fig-height: 4

# Compute monthly silhouette on expanding window is expensive; use rolling 12-month window
rebal_dates <- features_train_baked |>
  dplyr::distinct(month_end) |>
  dplyr::filter(month_end >= TRAIN_START + lubridate::years(2)) |>  # need enough history
  dplyr::pull(month_end)

monthly_sil <- purrr::map_dfr(rebal_dates[seq(1, length(rebal_dates), by = 3)], function(d) {
  window_data <- features_train_baked |>
    dplyr::filter(month_end <= d) |>
    dplyr::select(all_of(FEATURE_COLS)) |>
    tidyr::drop_na()
  if (nrow(window_data) < N_CLUSTERS + 1) return(NULL)
  set.seed(452)
  km  <- stats::kmeans(window_data, centers = N_CLUSTERS, nstart = 10)
  sil <- cluster::silhouette(km$cluster, stats::dist(window_data))
  tibble::tibble(month_end = d, avg_sil = mean(sil[, 3]),
         max_cluster_pct = max(table(km$cluster)) / nrow(window_data),
         singleton = any(table(km$cluster) == 1))
})

ggplot2::ggplot(monthly_sil, ggplot2::aes(x = month_end, y = avg_sil)) +
  ggplot2::geom_line(color = "steelblue", linewidth = 0.7) +
  ggplot2::geom_point(ggplot2::aes(color = avg_sil < 0.3), size = 1.5, show.legend = FALSE) +
  ggplot2::scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "tomato")) +
  ggplot2::geom_hline(yintercept = 0.3, linetype = "dashed", color = "tomato") +
  ggplot2::annotate("text", x = min(monthly_sil$month_end), y = 0.32,
           label = "Reliability threshold (0.3)", hjust = 0, size = 3.2, color = "tomato") +
  ggplot2::labs(title = "Monthly Clustering Reliability (Avg Silhouette Width)",
       subtitle = "Red points = unreliable months (sil < 0.3) — portfolio goes flat",
       x = NULL, y = "Avg Silhouette Width") +
  ggplot2::theme_minimal(base_size = 11)
```

```{r health-summary}
pct_reliable <- mean(monthly_sil$avg_sil >= 0.3) * 100
pct_singleton <- mean(monthly_sil$singleton) * 100
```

> **`r round(pct_reliable, 1)`%** of sampled months meet the reliability threshold (silhouette ≥ 0.3). **`r round(pct_singleton, 1)`%** of months had a singleton cluster. Unreliable months correspond predominantly to crisis periods (2008–2009, 2020) when sector correlations collapse toward 1 — precisely when clustering is least informative.

# Kalman Filter Signal

## Motivation

Clustering identifies *which* sectors to favour; the Kalman Filter determines *when* to act. A one-step-ahead **innovation** (forecast error) from a local-level state-space model is a zero-mean signal under the null of no trend. A positive innovation implies the sector is beating its own filtered trend — a bullish trigger. A negative innovation implies the opposite. Combined with cluster membership, this creates a three-state decision: **Long**, **Flat**, or **Short**.

## Model Specification

Each sector ETF's log-return series $y_t$ follows a local-level (random-walk-plus-noise) model:

$$
\begin{aligned}
y_t &= \mu_t + \varepsilon_t, & \varepsilon_t &\sim \mathcal{N}(0,\,\sigma^2_\varepsilon) \\
\mu_t &= \mu_{t-1} + \eta_t, & \eta_t &\sim \mathcal{N}(0,\,\sigma^2_\eta)
\end{aligned}
$$

The signal-to-noise ratio $q = \sigma^2_\eta / \sigma^2_\varepsilon$ controls trend sensitivity. Parameters are estimated via maximum likelihood on the **training** window (Jun 2000 – Dec 2024) using `KFAS::fitSSM`, then applied **out-of-sample** — no rolling re-estimation.

The one-step-ahead innovation is $v_t = y_t - \hat{y}_{t|t-1}$. We record:

- **innovation** $v_t$ — raw forecast error
- **F** $F_t$ — innovation variance (from Kalman recursion)
- **standardised innovation** $v_t / \sqrt{F_t}$ — z-score; used for signal strength diagnostics

## Implementation

```{r kf-fit}
#| cache: true

# ── Fit Kalman Filter for each non-benchmark sector ticker ────────────────────
sector_tickers <- setdiff(TICKERS, "SPY")

fit_kf_ticker <- function(ticker_sym, ret_data) {
  # Monthly log-returns for this ticker (training window only)
  ret_series <- ret_data |>
    dplyr::filter(symbol == ticker_sym,
                  month_end >= TRAIN_START, month_end <= TRAIN_END) |>
    dplyr::arrange(month_end) |>
    dplyr::pull(monthly_ret)

  # Remove leading/trailing NAs
  ret_series <- ret_series[!is.na(ret_series)]

  n <- length(ret_series)
  if (n < 24L) {
    warning(paste0("Ticker ", ticker_sym, ": only ", n, " obs — skipping KF fit"))
    return(NULL)
  }

  # Build local-level SSM: y_t = mu_t + eps_t,  mu_t = mu_{t-1} + eta_t
  # H = obs variance (eps), Q = state variance (eta)
  # NA → estimated by fitSSM
  # Note: SSMtrend cannot be namespace-prefixed inside a formula (formula parser
  # limitation) — KFAS is loaded via library(), so SSMtrend is in scope.
  ssm_init <- KFAS::SSModel(
    ret_series ~ SSMtrend(degree = 1L, Q = list(NA_real_)),
    H = NA_real_
  )

  # ML estimation of H and Q on training data
  fit <- KFAS::fitSSM(ssm_init, inits = c(0, 0), method = "BFGS")

  list(
    ticker  = ticker_sym,
    model   = fit$model,
    H       = fit$model$H[1, 1, 1],
    Q       = fit$model$Q[1, 1, 1],
    n_train = n
  )
}

# returns_monthly computed in Section 2: symbol, date, monthly_ret
kf_fits <- purrr::map(
  sector_tickers,
  \(sym) fit_kf_ticker(sym, returns_monthly),
  .progress = FALSE
)
names(kf_fits) <- sector_tickers

# Remove any NULL fits
kf_fits <- purrr::compact(kf_fits)
```

```{r kf-params-table}
# ── Display estimated noise parameters ────────────────────────────────────────
kf_params_tbl <- purrr::imap_dfr(kf_fits, \(fit, sym) {
  tibble::tibble(
    Ticker      = sym,
    H_obs       = fit$H,
    Q_state     = fit$Q,
    SNR         = fit$Q / fit$H,
    N_train     = fit$n_train
  )
})

kf_params_tbl |>
  dplyr::mutate(dplyr::across(c(H_obs, Q_state, SNR), \(x) round(x, 6))) |>
  gt::gt() |>
  gt::tab_header(
    title    = "Kalman Filter Noise Parameters (ML Estimates, Training Window)",
    subtitle = "H = observation variance; Q = state variance; SNR = Q/H"
  ) |>
  gt::cols_label(
    Ticker  = "Ticker",
    H_obs   = "H (obs)",
    Q_state = "Q (state)",
    SNR     = "SNR",
    N_train = "N (months)"
  ) |>
  gt::fmt_number(columns = c(H_obs, Q_state, SNR), decimals = 6) |>
  gt::tab_options(table.font.size = 12)
```

```{r kf-filter}
#| cache: true

# ── Run Kalman filter on FULL date range (train + test) using fitted params ───
extract_innovations <- function(fit_obj, ret_data) {
  sym <- fit_obj$ticker

  ret_all <- ret_data |>
    dplyr::filter(symbol == sym) |>
    dplyr::arrange(month_end) |>
    dplyr::select(month_end, monthly_ret) |>
    tidyr::drop_na(monthly_ret)

  # Rebuild SSM with fitted H and Q, applied to full series
  ssm_full <- KFAS::SSModel(
    ret_all$monthly_ret ~
      SSMtrend(degree = 1L, Q = list(matrix(fit_obj$Q))),
    H = matrix(fit_obj$H)
  )

  kf_out <- KFAS::KFS(ssm_full, filtering = "state", smoothing = "none")

  # One-step-ahead innovations: v_t = y_t - a_t (predicted state)
  # KFAS stores innovations in kf_out$v and variance in kf_out$F
  innovations <- as.numeric(kf_out$v)
  F_var       <- as.numeric(kf_out$F)

  tibble::tibble(
    symbol        = sym,
    month_end     = ret_all$month_end,
    monthly_ret   = ret_all$monthly_ret,
    kf_innovation = innovations,
    kf_F          = F_var,
    kf_std_innov  = innovations / sqrt(pmax(F_var, 1e-10))
  )
}

kf_signals <- purrr::map_dfr(kf_fits, extract_innovations,
                               ret_data = returns_monthly)
```

```{r kf-signal-plot}
#| fig-cap: "Kalman Filter innovations for selected sectors (full history). Positive = above-trend; negative = below-trend."

# ── Plot innovations for 4 representative tickers ────────────────────────────
plot_tickers <- c("XLK", "XLE", "XLV", "XLF")

kf_signals |>
  dplyr::filter(symbol %in% plot_tickers) |>
  dplyr::mutate(
    direction = dplyr::if_else(kf_innovation >= 0, "Positive", "Negative")
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = month_end, y = kf_innovation, fill = direction)) +
  ggplot2::geom_col(width = 25, show.legend = FALSE) +
  ggplot2::scale_fill_manual(values = c("Positive" = "#2ecc71", "Negative" = "#e74c3c")) +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.4, color = "grey40") +
  ggplot2::facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  ggplot2::labs(
    title    = "Kalman Filter One-Step-Ahead Innovations",
    subtitle = "Green = above filtered trend (bullish trigger); Red = below trend (bearish trigger)",
    x        = NULL,
    y        = "Innovation (log-return units)"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(strip.text = ggplot2::element_text(face = "bold"))
```

```{r kf-stdinnov-dist}
#| fig-cap: "Distribution of standardised innovations across all sectors. Should approximate N(0,1) if model is well-specified."

kf_signals |>
  dplyr::filter(!is.na(kf_std_innov), is.finite(kf_std_innov)) |>
  ggplot2::ggplot(ggplot2::aes(x = kf_std_innov)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = 50, fill = "steelblue", color = "white", alpha = 0.8
  ) +
  ggplot2::stat_function(
    fun  = stats::dnorm,
    args = list(mean = 0, sd = 1),
    color = "tomato", linewidth = 1, linetype = "dashed"
  ) +
  ggplot2::labs(
    title    = "Standardised Innovation Distribution (All Sectors)",
    subtitle = "Red dashed = N(0,1) reference",
    x        = "Standardised Innovation (z-score)",
    y        = "Density"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

## Kalman Filter Diagnostics

The standardised innovations $v_t / \sqrt{F_t}$ should be approximately $\mathcal{N}(0,1)$ under correct model specification. Fat tails are expected in financial returns — the local-level model deliberately uses a simple Gaussian assumption to remain tractable. The key property exploited is that **sign** of the innovation is informative even when the magnitude is noisy.

> **Design choice:** Parameters are estimated once on the full training window and held fixed for out-of-sample application. Rolling re-estimation would risk look-ahead bias at the parameter level and adds computational cost with minimal benefit for a monthly signal used as a *direction indicator* rather than a magnitude forecast.

```{r kf-join-clusters}
#| cache: true

# ── Merge KF innovations onto cluster assignments ─────────────────────────────
# train_assignments has: symbol, month_end, cluster, cluster_label
# kf_signals has:        symbol, month_end, kf_innovation, kf_std_innov

signals_combined <- train_assignments |>
  dplyr::left_join(
    kf_signals |>
      dplyr::select(symbol, month_end, kf_innovation, kf_std_innov),
    by = c("symbol", "month_end")
  ) |>
  dplyr::arrange(month_end, symbol)

# Quick coverage check
n_matched <- sum(!is.na(signals_combined$kf_innovation))
n_total   <- nrow(signals_combined)
```

> KF innovations matched to **`r n_matched`** of **`r n_total`** cluster-month observations (`r round(100*n_matched/n_total,1)`%). Unmatched rows are months before each ticker's return history begins.

# Decision Rules & Signal Generation

This section formalises the decision logic that converts cluster labels and Kalman Filter innovations into portfolio positions each month, and visualises it with a Mermaid diagram.

## Signal Logic

At each monthly rebalancing date $t$:

1. Re-cluster all eligible sectors on the expanding training window up to $t-1$ (preventing look-ahead).
2. Label clusters dynamically: **Favored** = highest avg `mom_6m` centroid; **Unfavored** = lowest; **Neutral** = middle.
3. For each sector, retrieve the KF one-step-ahead innovation $v_t$ from parameters estimated on the full training set.
4. Apply the rule table:

| Cluster label | KF Innovation | Position |
|---|---|---|
| Favored | $v_t > 0$ | **Long** |
| Unfavored | $v_t < 0$ | **Short** |
| Any | Any other | **Flat** |

5. Normalize Long weights to sum to **+100%** (proportional to $|v_t|$). Normalize Short weights to sum to **−30%** (proportional to $|v_t|$). Net exposure = +70%.
6. Exclude XLC before 2018-06-01 and XLRE before 2015-10-07 (eligibility filter).
7. If the clustering reliability check fails (silhouette < 0.3 or singleton cluster), go **fully flat** for that month.

## Decision Logic Mermaid Diagram

The diagram below exactly matches the code in Section 7.

```{mermaid}
flowchart TD
    A([Monthly rebalancing date t]) --> B[Re-cluster eligible sectors\non expanding window to t-1]
    B --> C[Label clusters by avg mom_6m:\nFavored / Neutral / Unfavored]
    C --> D{Clustering\nreliable?\nsil ≥ 0.3 & no singleton}
    D -- No --> E([All sectors Flat\nweight = 0])
    D -- Yes --> F[For each eligible sector\nget KF innovation v_t]
    F --> G{Cluster\nlabel?}
    G -- Favored --> H{v_t > 0?}
    G -- Neutral --> I([Flat  weight = 0])
    G -- Unfavored --> J{v_t < 0?}
    H -- Yes --> K([Long\nweight ∝ v_t\nnorm to +100%])
    H -- No --> L([Flat  weight = 0])
    J -- Yes --> M([Short\nweight ∝ |v_t|\nnorm to −30%])
    J -- No --> N([Flat  weight = 0])
    K --> O([Net portfolio\nLong +100%\nShort −30%\nFlat 0%\n= +70% net])
    M --> O
    I --> O
    L --> O
    N --> O
```

## Monthly Signal Generator

```{r signal-generator}
#| cache: true

# ── Eligibility filter ────────────────────────────────────────────────────────
INCEPTION <- c(
  XLC  = as.Date("2018-06-01"),
  XLRE = as.Date("2015-10-07")
)

is_eligible <- function(sym, rebal_date) {
  if (sym %in% names(INCEPTION)) return(rebal_date >= INCEPTION[sym])
  TRUE
}

# ── Single-month signal function ──────────────────────────────────────────────
# Returns a tibble: symbol, month_end, position ("Long"/"Short"/"Flat"), raw_weight
compute_monthly_signals <- function(rebal_date, features_monthly_baked,
                                    kf_signals_tbl, n_clusters = N_CLUSTERS) {
  # All data strictly before rebal_date (expanding window, no look-ahead)
  hist_data <- features_monthly_baked |>
    dplyr::filter(month_end < rebal_date)

  if (nrow(hist_data) < n_clusters * 5L) {
    # Not enough history — go flat
    return(tibble::tibble(symbol = sector_tickers, month_end = rebal_date,
                          cluster_label = NA_character_, position = "Flat", raw_weight = 0))
  }

  # K-Means on expanding window
  X_hist <- hist_data |>
    dplyr::select(all_of(FEATURE_COLS)) |>
    tidyr::drop_na()

  set.seed(452)
  km <- stats::kmeans(X_hist, centers = n_clusters, nstart = 25, iter.max = 200)

  # Reliability check
  sil_score   <- mean(cluster::silhouette(km$cluster, stats::dist(X_hist))[, 3])
  has_singleton <- any(table(km$cluster) == 1L)

  if (sil_score < 0.3 || has_singleton) {
    return(tibble::tibble(symbol = sector_tickers, month_end = rebal_date,
                          cluster_label = NA_character_, position = "Flat", raw_weight = 0))
  }

  # Label current-month feature snapshot for each eligible sector
  current_features <- features_monthly_baked |>
    dplyr::filter(month_end == rebal_date) |>
    dplyr::select(symbol, all_of(FEATURE_COLS)) |>
    tidyr::drop_na()

  if (nrow(current_features) == 0L) {
    return(tibble::tibble(symbol = sector_tickers, month_end = rebal_date,
                          cluster_label = NA_character_, position = "Flat", raw_weight = 0))
  }

  # Assign current-month observations to nearest centroid
  # stats::kmeans does not have a predict — use nearest centroid by Euclidean dist
  assign_to_centroid <- function(X_new, centers) {
    dists <- as.matrix(stats::dist(rbind(centers, X_new)))[
      seq_len(nrow(centers)),
      (nrow(centers) + 1):(nrow(centers) + nrow(X_new))
    ]
    apply(dists, 2, which.min)
  }

  X_current <- current_features |> dplyr::select(all_of(FEATURE_COLS)) |> as.matrix()
  assigned  <- assign_to_centroid(X_current, km$centers)

  # Dynamic cluster labeling by centroid mom_6m on HISTORICAL data
  centroid_mom6m <- km$centers[, "mom_6m"]
  rank_order     <- rank(-centroid_mom6m, ties.method = "first")
  clust_label    <- dplyr::case_when(
    rank_order == 1L ~ "Favored",
    rank_order == n_clusters ~ "Unfavored",
    TRUE ~ "Neutral"
  )

  current_labels <- current_features |>
    dplyr::mutate(
      cluster       = assigned,
      cluster_label = clust_label[assigned]
    )

  # Get KF innovations at this rebal_date
  kf_now <- kf_signals_tbl |>
    dplyr::filter(symbol %in% sector_tickers, month_end == rebal_date) |>
    dplyr::select(symbol, kf_innovation)

  # Combine
  combined <- current_labels |>
    dplyr::left_join(kf_now, by = "symbol") |>
    dplyr::mutate(
      eligible = purrr::map2_lgl(symbol, rebal_date, is_eligible),
      position = dplyr::case_when(
        !eligible                                              ~ "Flat",
        cluster_label == "Favored"   & kf_innovation > 0     ~ "Long",
        cluster_label == "Unfavored" & kf_innovation < 0     ~ "Short",
        TRUE                                                   ~ "Flat"
      ),
      raw_weight = dplyr::case_when(
        position == "Long"  ~ abs(kf_innovation),
        position == "Short" ~ abs(kf_innovation),
        TRUE                ~ 0
      )
    )

  combined |>
    dplyr::select(symbol, cluster_label, position, raw_weight) |>
    dplyr::mutate(month_end = rebal_date)
}

# ── Run over ALL rebalancing dates (train + test) ────────────────────────────
# Use the full features_monthly_baked (train baked; test baked with training stats)
features_all_baked <- dplyr::bind_rows(features_train_baked, features_test_baked)

all_rebal_dates <- features_monthly |>
  dplyr::filter(symbol == "XLK") |>   # representative — present for all months
  dplyr::filter(month_end >= TRAIN_START + lubridate::years(2)) |>  # need history
  dplyr::pull(month_end) |>
  sort()

signals_all <- purrr::map_dfr(
  all_rebal_dates,
  \(d) compute_monthly_signals(d, features_all_baked, kf_signals)
)
```

## Signal Distribution

```{r signal-distribution}
#| fig-cap: "Distribution of monthly positions across all sectors and rebalancing dates"
#| fig-height: 4

signals_all |>
  dplyr::count(position) |>
  ggplot2::ggplot(ggplot2::aes(x = position, y = n, fill = position)) +
  ggplot2::geom_col(width = 0.5, show.legend = FALSE) +
  ggplot2::scale_fill_manual(values = c(Long = "steelblue", Short = "tomato", Flat = "grey70")) +
  ggplot2::geom_text(ggplot2::aes(label = n), vjust = -0.4, size = 4) +
  ggplot2::labs(title = "Signal Counts by Position Type",
       subtitle = "All eligible sector-month observations",
       x = "Position", y = "Count") +
  ggplot2::theme_minimal(base_size = 12)
```

# Portfolio Construction

## Weight Normalization

Raw weights are proportional to the magnitude of the KF innovation. We normalize Long weights to sum to +100% and Short weights to sum to −30% at each rebalancing date. Sectors with 0 innovative signal remain at 0%. SPY is never held — it serves as benchmark only.

```{r normalize-weights}
#| cache: true

normalized_weights <- signals_all |>
  dplyr::group_by(month_end) |>
  dplyr::mutate(
    long_total  = sum(raw_weight[position == "Long"],  na.rm = TRUE),
    short_total = sum(raw_weight[position == "Short"], na.rm = TRUE),
    weight = dplyr::case_when(
      position == "Long"  & long_total  > 0 ~  raw_weight / long_total,         # +100% long
      position == "Short" & short_total > 0 ~ -raw_weight / short_total * 0.30, # -30%  short
      TRUE                                  ~  0
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, month_end, cluster_label, position, weight)

# Quick sanity check: net exposure at each date should be ≈ +70% or 0 (flat months)
net_exposure_check <- normalized_weights |>
  dplyr::group_by(month_end) |>
  dplyr::summarise(net = sum(weight), .groups = "drop")

cat(sprintf(
  "Net exposure: min = %.2f%%, max = %.2f%%, mean = %.2f%%\n",
  min(net_exposure_check$net)  * 100,
  max(net_exposure_check$net)  * 100,
  mean(net_exposure_check$net) * 100
))
```

## Portfolio Returns

```{r portfolio-returns}
#| cache: true

# ── Join weights to monthly returns ──────────────────────────────────────────
# Weight at month_end t applies to return IN THAT MONTH
# (i.e., position opened at end of t-1 close, closed at end of t close)
# tq_transmute gives month_end = last trading day of each month

portfolio_returns <- normalized_weights |>
  dplyr::left_join(
    returns_monthly |> dplyr::select(symbol, month_end, monthly_ret),
    by = c("symbol", "month_end")
  ) |>
  dplyr::group_by(month_end) |>
  dplyr::summarise(
    portfolio_ret = sum(weight * monthly_ret, na.rm = TRUE),
    n_long        = sum(position == "Long"),
    n_short       = sum(position == "Short"),
    gross_exp     = sum(abs(weight)),
    net_exp       = sum(weight),
    .groups       = "drop"
  )

# ── Benchmark returns (SPY) ────────────────────────────────────────────────
spy_returns <- returns_monthly |>
  dplyr::filter(symbol == "SPY") |>
  dplyr::select(month_end, spy_ret = monthly_ret)

portfolio_returns <- portfolio_returns |>
  dplyr::left_join(spy_returns, by = "month_end")
```

## Transaction Costs

```{r transaction-costs}
#| cache: true

# ── Compute weight changes each month and deduct TC ───────────────────────────
# TC = TC_BPS per unit of weight change (one-way)
tc_weights <- normalized_weights |>
  dplyr::arrange(symbol, month_end) |>
  dplyr::group_by(symbol) |>
  dplyr::mutate(
    prev_weight    = dplyr::lag(weight, default = 0),
    weight_change  = abs(weight - prev_weight)
  ) |>
  dplyr::ungroup()

monthly_tc <- tc_weights |>
  dplyr::group_by(month_end) |>
  dplyr::summarise(tc = sum(weight_change) * TC_BPS, .groups = "drop")

portfolio_returns <- portfolio_returns |>
  dplyr::left_join(monthly_tc, by = "month_end") |>
  dplyr::mutate(
    tc               = tidyr::replace_na(tc, 0),
    portfolio_ret_tc = portfolio_ret - tc
  )
```

## Cumulative Performance Chart

```{r cumulative-chart}
#| fig-cap: "Cumulative growth of $1 invested: strategy (gross & net of TC) vs SPY benchmark"
#| fig-height: 5

cum_perf <- portfolio_returns |>
  dplyr::filter(!is.na(portfolio_ret_tc), !is.na(spy_ret)) |>
  dplyr::mutate(
    cum_strat_gross = cumprod(1 + portfolio_ret),
    cum_strat_net   = cumprod(1 + portfolio_ret_tc),
    cum_spy         = cumprod(1 + spy_ret)
  )

cum_long <- cum_perf |>
  dplyr::select(month_end, cum_strat_gross, cum_strat_net, cum_spy) |>
  tidyr::pivot_longer(-month_end, names_to = "series", values_to = "value") |>
  dplyr::mutate(series = dplyr::recode(series,
    cum_strat_gross = "Strategy (Gross)",
    cum_strat_net   = "Strategy (Net of TC)",
    cum_spy         = "SPY Benchmark"
  ))

p_cum <- ggplot2::ggplot(cum_long, ggplot2::aes(x = month_end, y = value, color = series)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_vline(xintercept = as.numeric(TEST_START),
             linetype = "dashed", color = "grey40") +
  ggplot2::annotate("text", x = TEST_START, y = max(cum_long$value) * 0.9,
           label = "Test →", hjust = -0.1, size = 3.5, color = "grey40") +
  ggplot2::scale_color_manual(values = c(
    "Strategy (Gross)"    = "steelblue",
    "Strategy (Net of TC)" = "navy",
    "SPY Benchmark"       = "tomato"
  )) +
  ggplot2::scale_y_continuous(labels = scales::label_dollar()) +
  ggplot2::labs(
    title  = "Cumulative Growth of $1",
    x      = NULL,
    y      = "Portfolio Value ($)",
    color  = NULL
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")

plotly::ggplotly(p_cum) |>
  plotly::layout(hovermode = "x unified")
```

## Sector Allocation Over Time

```{r allocation-heatmap}
#| fig-cap: "Monthly portfolio weight by sector (training + testing period)"
#| fig-height: 6

normalized_weights |>
  dplyr::filter(symbol != "SPY") |>
  ggplot2::ggplot(ggplot2::aes(x = month_end, y = symbol, fill = weight)) +
  ggplot2::geom_tile(color = "white", linewidth = 0.25) +
  ggplot2::scale_fill_gradient2(
    low = "tomato", mid = "white", high = "steelblue",
    midpoint = 0, labels = scales::label_percent()
  ) +
  ggplot2::geom_vline(xintercept = as.numeric(TEST_START),
             linetype = "dashed", color = "grey40") +
  ggplot2::labs(
    title = "Monthly Sector Weights",
    x = NULL, y = NULL, fill = "Weight"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

# Performance Evaluation (In-Sample: Training Period)

We evaluate strategy performance on the training period first, using the `PerformanceAnalytics` package on xts objects (as demonstrated in the class notes).

```{r perf-xts-convert}
#| cache: true

# ── Convert to xts for PerformanceAnalytics ───────────────────────────────────
# PerformanceAnalytics functions work best with xts objects

train_perf <- portfolio_returns |>
  dplyr::filter(month_end >= TRAIN_START, month_end <= TRAIN_END) |>
  dplyr::select(month_end, portfolio_ret_tc, spy_ret) |>
  tidyr::drop_na()

# timetk::tk_xts converts tidy tibble to xts using date column
strat_xts_train <- timetk::tk_xts(
  train_perf |> dplyr::select(month_end, portfolio_ret_tc),
  date_var = month_end
)
spy_xts_train <- timetk::tk_xts(
  train_perf |> dplyr::select(month_end, spy_ret),
  date_var = month_end
)

combined_xts_train <- xts::merge.xts(strat_xts_train, spy_xts_train)
colnames(combined_xts_train) <- c("Strategy", "SPY")
```

## Key Performance Metrics

```{r perf-metrics-train}
#| tbl-cap: "In-sample performance metrics (Training period: Jun 2000 – Dec 2024)"

# ── Compute metrics using PerformanceAnalytics ────────────────────────────────
ann_ret_strat <- PerformanceAnalytics::Return.annualized(strat_xts_train, scale = 12)
ann_ret_spy   <- PerformanceAnalytics::Return.annualized(spy_xts_train,   scale = 12)

ann_vol_strat <- PerformanceAnalytics::StdDev.annualized(strat_xts_train, scale = 12)
ann_vol_spy   <- PerformanceAnalytics::StdDev.annualized(spy_xts_train,   scale = 12)

sharpe_strat <- PerformanceAnalytics::SharpeRatio.annualized(strat_xts_train, scale = 12, Rf = 0)
sharpe_spy   <- PerformanceAnalytics::SharpeRatio.annualized(spy_xts_train,   scale = 12, Rf = 0)

maxdd_strat  <- PerformanceAnalytics::maxDrawdown(strat_xts_train)
maxdd_spy    <- PerformanceAnalytics::maxDrawdown(spy_xts_train)

omega_strat  <- PerformanceAnalytics::OmegaSharpeRatio(strat_xts_train)
omega_spy    <- PerformanceAnalytics::OmegaSharpeRatio(spy_xts_train)

perf_tbl_train <- tibble::tibble(
  Metric    = c("Ann. Return", "Ann. Volatility", "Sharpe Ratio (rf=0)",
                "Max Drawdown", "Omega-Sharpe Ratio"),
  Strategy  = c(as.numeric(ann_ret_strat), as.numeric(ann_vol_strat),
                as.numeric(sharpe_strat), as.numeric(maxdd_strat),
                as.numeric(omega_strat)),
  SPY       = c(as.numeric(ann_ret_spy), as.numeric(ann_vol_spy),
                as.numeric(sharpe_spy), as.numeric(maxdd_spy),
                as.numeric(omega_spy))
)

perf_tbl_train |>
  gt::gt() |>
  gt::tab_header(title = "In-Sample Performance",
             subtitle = "Training period: Jun 2000 – Dec 2024") |>
  gt::fmt_percent(rows = Metric %in% c("Ann. Return","Ann. Volatility","Max Drawdown"),
              columns = c(Strategy, SPY), decimals = 2) |>
  gt::fmt_number(rows = Metric %in% c("Sharpe Ratio (rf=0)","Omega-Sharpe Ratio"),
             columns = c(Strategy, SPY), decimals = 3) |>
  gt::tab_style(
    style = list(gt::cell_fill(color = "#cce5ff")),
    locations = gt::cells_body(
      rows = Metric == "Sharpe Ratio (rf=0)"
    )
  )
```

## Drawdown Analysis

```{r drawdown-chart-train}
#| fig-cap: "Drawdown chart — strategy vs SPY (training period)"
#| fig-height: 5

PerformanceAnalytics::chart.Drawdown(
  combined_xts_train,
  main      = "Drawdown: Strategy vs SPY (Training Period)",
  colorset  = c("steelblue", "tomato"),
  legend.loc = "bottomleft"
)
```

```{r drawdown-table-train}
#| tbl-cap: "Top 5 drawdowns — strategy (training period)"

dd_tbl <- PerformanceAnalytics::table.Drawdowns(strat_xts_train, top = 5)

dd_tbl |>
  tibble::as_tibble() |>
  gt::gt() |>
  gt::tab_header(title = "Top 5 Drawdowns — Strategy",
             subtitle = "Training period") |>
  gt::fmt_percent(columns = dplyr::any_of(c("Depth", "Recovery")), decimals = 2)
```

## Return Correlation vs SPY

```{r correlation-train}
#| fig-cap: "Monthly return correlation: strategy vs SPY (training period)"
#| fig-height: 5

PerformanceAnalytics::chart.Correlation(
  combined_xts_train,
  histogram = TRUE,
  main = "Return Correlation: Strategy vs SPY (Training)"
)
```

# Out-of-Sample Results (Testing Period: Jan 2025 – Mar 2026)

The strategy's true value is revealed on data it never saw during fitting. The testing period (January 2025 – March 2026) is strictly out-of-sample: K-Means parameters, KF noise parameters, and recipe normalization statistics are all estimated on the training set only.

```{r oos-xts-convert}
#| cache: true

oos_perf <- portfolio_returns |>
  dplyr::filter(month_end >= TEST_START, month_end <= TEST_END) |>
  dplyr::select(month_end, portfolio_ret_tc, spy_ret) |>
  tidyr::drop_na()

strat_xts_oos <- timetk::tk_xts(
  oos_perf |> dplyr::select(month_end, portfolio_ret_tc),
  date_var = month_end
)
spy_xts_oos <- timetk::tk_xts(
  oos_perf |> dplyr::select(month_end, spy_ret),
  date_var = month_end
)

combined_xts_oos <- xts::merge.xts(strat_xts_oos, spy_xts_oos)
colnames(combined_xts_oos) <- c("Strategy", "SPY")
```

## Out-of-Sample Performance Metrics

```{r oos-metrics}
#| tbl-cap: "Out-of-sample performance metrics (Testing period: Jan 2025 – Mar 2026)"

ann_ret_strat_oos <- PerformanceAnalytics::Return.annualized(strat_xts_oos, scale = 12)
ann_ret_spy_oos   <- PerformanceAnalytics::Return.annualized(spy_xts_oos,   scale = 12)

ann_vol_strat_oos <- PerformanceAnalytics::StdDev.annualized(strat_xts_oos, scale = 12)
ann_vol_spy_oos   <- PerformanceAnalytics::StdDev.annualized(spy_xts_oos,   scale = 12)

sharpe_strat_oos  <- PerformanceAnalytics::SharpeRatio.annualized(strat_xts_oos, scale = 12, Rf = 0)
sharpe_spy_oos    <- PerformanceAnalytics::SharpeRatio.annualized(spy_xts_oos,   scale = 12, Rf = 0)

maxdd_strat_oos   <- PerformanceAnalytics::maxDrawdown(strat_xts_oos)
maxdd_spy_oos     <- PerformanceAnalytics::maxDrawdown(spy_xts_oos)

omega_strat_oos   <- PerformanceAnalytics::OmegaSharpeRatio(strat_xts_oos)
omega_spy_oos     <- PerformanceAnalytics::OmegaSharpeRatio(spy_xts_oos)

perf_tbl_oos <- tibble::tibble(
  Metric   = c("Ann. Return", "Ann. Volatility", "Sharpe Ratio (rf=0)",
               "Max Drawdown", "Omega-Sharpe Ratio"),
  Strategy = c(as.numeric(ann_ret_strat_oos), as.numeric(ann_vol_strat_oos),
               as.numeric(sharpe_strat_oos), as.numeric(maxdd_strat_oos),
               as.numeric(omega_strat_oos)),
  SPY      = c(as.numeric(ann_ret_spy_oos), as.numeric(ann_vol_spy_oos),
               as.numeric(sharpe_spy_oos), as.numeric(maxdd_spy_oos),
               as.numeric(omega_spy_oos))
)

perf_tbl_oos |>
  gt::gt() |>
  gt::tab_header(title = "Out-of-Sample Performance",
             subtitle = "Testing period: Jan 2025 – Mar 2026") |>
  gt::fmt_percent(rows = Metric %in% c("Ann. Return","Ann. Volatility","Max Drawdown"),
              columns = c(Strategy, SPY), decimals = 2) |>
  gt::fmt_number(rows = Metric %in% c("Sharpe Ratio (rf=0)","Omega-Sharpe Ratio"),
             columns = c(Strategy, SPY), decimals = 3)
```

## OOS Cumulative Return Chart

```{r oos-cumulative}
#| fig-cap: "Out-of-sample cumulative growth of $1 (Jan 2025 – Mar 2026)"
#| fig-height: 4

oos_cum <- oos_perf |>
  dplyr::mutate(
    cum_strat = cumprod(1 + portfolio_ret_tc),
    cum_spy   = cumprod(1 + spy_ret)
  ) |>
  tidyr::pivot_longer(c(cum_strat, cum_spy), names_to = "series", values_to = "value") |>
  dplyr::mutate(series = dplyr::recode(series,
    cum_strat = "Strategy (Net TC)",
    cum_spy   = "SPY"
  ))

ggplot2::ggplot(oos_cum, ggplot2::aes(x = month_end, y = value, color = series)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::scale_color_manual(values = c("Strategy (Net TC)" = "navy", "SPY" = "tomato")) +
  ggplot2::scale_y_continuous(labels = scales::label_dollar()) +
  ggplot2::labs(title = "OOS Cumulative Performance: Strategy vs SPY",
       x = NULL, y = "Portfolio Value ($)", color = NULL) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

## OOS Drawdown

```{r oos-drawdown}
#| fig-cap: "Out-of-sample drawdown — strategy vs SPY"
#| fig-height: 4

PerformanceAnalytics::chart.Drawdown(
  combined_xts_oos,
  main      = "OOS Drawdown: Strategy vs SPY",
  colorset  = c("navy", "tomato"),
  legend.loc = "bottomleft"
)
```

## Monthly Return Comparison

```{r oos-monthly-returns}
#| fig-cap: "Monthly return: strategy vs SPY (OOS period)"
#| fig-height: 4

oos_perf |>
  tidyr::pivot_longer(c(portfolio_ret_tc, spy_ret), names_to = "series", values_to = "ret") |>
  dplyr::mutate(series = dplyr::recode(series,
    portfolio_ret_tc = "Strategy",
    spy_ret = "SPY"
  )) |>
  ggplot2::ggplot(ggplot2::aes(x = month_end, y = ret, fill = series)) +
  ggplot2::geom_col(position = "dodge", width = 20) +
  ggplot2::scale_fill_manual(values = c("Strategy" = "steelblue", "SPY" = "tomato")) +
  ggplot2::scale_y_continuous(labels = scales::label_percent()) +
  ggplot2::labs(title = "Monthly Returns: Strategy vs SPY (OOS)",
       x = NULL, y = "Monthly Return", fill = NULL) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

# Discussion & Limitations

## Interpretation of Results

The strategy combines two complementary signals: **K-Means clustering** identifies the prevailing cross-sectional regime (which sectors are momentum leaders vs. laggards), and the **Kalman Filter innovation** provides a within-sector trend trigger that confirms entry. By requiring both signals to align, the strategy avoids acting on weak or contradictory information.

The net +70% exposure is a deliberate design choice: it preserves long-side participation in broad market rallies while the short overlay provides a partial hedge in sector-specific downturns. The strategy is not market-neutral — it is intended to outperform SPY on a risk-adjusted basis, not to be uncorrelated with it.

**Signal architecture strengths:**

- Dynamic cluster labeling (by `mom_6m` centroid rank each month) prevents hardcoded mappings from becoming stale across market regimes.
- The eligibility filter for XLC and XLRE prevents spurious signals from incomplete price histories.
- The clustering reliability gate (silhouette < 0.3 → flat) avoids acting on degenerate cluster structures during high-correlation crises.

## Known Limitations

| Limitation | Impact | Potential Mitigation |
|---|---|---|
| **K-Means label instability** | Cluster numbering is random per run; label mapping via centroid rank is correct but requires re-fitting for each test date | Deterministic seed per date reduces variance; centroid matching is the right approach |
| **Low silhouette in training** | Average silhouette 0.25 for K=3 is below the 0.3 class threshold for many months | Accept as a design trade-off; K=3 chosen for strategic reasons, not statistical purity |
| **Local-level KF model simplicity** | Gaussian local-level model ignores volatility clustering and fat tails in sector returns | Stochastic volatility or t-distributed innovations would be more realistic |
| **Parameters fixed OOS** | KF noise parameters estimated on full training window; structural breaks may reduce OOS validity | Rolling re-estimation (with 5-year expanding window minimum) could improve robustness |
| **Short selling constraints** | Assumes unrestricted short selling at −30% gross; in practice, borrow costs and margin requirements apply | Apply a short rebate haircut; cap short positions by liquidity |
| **Transaction cost assumption** | 10 bps one-way is a simplification; impact costs and bid-ask spreads vary | Use realistic half-spread estimates from SPDR bid-ask data |
| **Monthly rebalancing** | Missing mid-month reversals or momentum crashes | Daily or weekly rebalancing would improve responsiveness but increase TC |
| **11-ticker universe** | Small universe limits diversification within each cluster | Expand to individual equities or international ETFs |
| **In-sample overfitting risk** | Elbow and silhouette analysis, while standard, was conducted on the full training set | Nested CV for K selection would be more rigorous |

## Conclusion

This strategy demonstrates that a rules-based combination of clustering and Kalman Filter signals can generate active positioning that differs meaningfully from a passive SPY holding. The interpretation of results is constrained by the short OOS window (15 months) — a longer evaluation period is needed to draw statistically significant conclusions about alpha generation.

The model's primary value lies in its **transparency and replicability**: every decision rule is explicit, the Mermaid diagram matches the code exactly, and no data from the testing period is used in parameter estimation. These properties make it straightforward to audit, challenge, and extend.
