---
title: "Clustering-Based Sector Rotation with Kalman Filter Signals"
subtitle: "FIN 452 — Quantitative Trading Model"
author: "Nursultan Berdibek"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    fig-dpi: 150
    theme: cosmo
execute:
  cache: true
  message: false
  warning: false
  echo: true
---

```{r setup}
#| cache: false

# ── Packages ──────────────────────────────────────────────────────────────────
library(tidyverse)
library(tidyquant)
library(TTR)
library(tidymodels)
library(tidyclust)
library(factoextra)
library(KFAS)
library(PerformanceAnalytics)
library(plotly)
library(patchwork)
library(gt)
library(lubridate)
library(scales)

# ── Global Constants ──────────────────────────────────────────────────────────
TICKERS <- c("SPY", "XLB", "XLC", "XLE", "XLF",
             "XLI", "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")

SECTOR_NAMES <- c(
  SPY  = "S&P 500 (Benchmark)",
  XLB  = "Materials",
  XLC  = "Communication Services",
  XLE  = "Energy",
  XLF  = "Financials",
  XLI  = "Industrials",
  XLK  = "Technology",
  XLP  = "Consumer Staples",
  XLRE = "Real Estate",
  XLV  = "Health Care",
  XLU  = "Utilities",
  XLY  = "Consumer Discretionary"
)

TRAIN_START <- as.Date("2000-06-01")
TRAIN_END   <- as.Date("2024-12-31")
TEST_START  <- as.Date("2025-01-01")
TEST_END    <- as.Date("2026-03-31")

N_CLUSTERS       <- 3L       # justified via elbow plot in Section 4
MOMENTUM_WINDOW  <- 6L       # months for favored-cluster labeling
KF_INNOV_CLIP_SD <- 3.0      # clip KF innovations beyond ±3 SD
TC_BPS           <- 0.0010   # 10 bps one-way transaction cost

set.seed(452)
```

```{r pkg-versions}
#| cache: false
#| tbl-cap: "Package versions used in this analysis"

pkgs <- c("tidyverse", "tidyquant", "TTR", "tidymodels", "tidyclust",
          "factoextra", "KFAS", "PerformanceAnalytics", "plotly",
          "patchwork", "gt", "lubridate", "scales")

tibble(
  Package = pkgs,
  Version = map_chr(pkgs, \(p) as.character(packageVersion(p)))
) |>
  gt() |>
  tab_header(title = "R Package Versions") |>
  cols_align("left")
```

# Introduction & Strategy Overview

This report implements a **clustering-based sector rotation strategy** applied to the eleven S&P 500 Select Sector SPDR ETFs, using SPY as a benchmark. The strategy combines two complementary methods:

1. **K-Means Clustering** groups sectors by their return and momentum characteristics each month. The cluster with the strongest trailing momentum is designated the *favored* cluster — the universe eligible for investment.

2. **Kalman Filter** fits a local-level state-space model to each sector's return history. The one-step-ahead *innovation* (observed minus predicted) measures whether a sector is outperforming its recent trend. Only sectors in the favored cluster with a positive Kalman innovation receive a positive allocation.

Portfolio weights are normalized to sum to 100% (long-only). Monthly rebalancing occurs on the last trading day of each month, with execution assumed at the same day's closing price. A 10 bps one-way transaction cost is applied proportional to weight changes.

**Training period:** June 2000 – December 2024
**Testing period:** January 2025 – March 2026

# Data Acquisition & Validation

## Pulling Price Data

All data is sourced from Yahoo Finance via the `tidyquant` package. No local data files are used. Adjusted closing prices account for dividends and splits.

```{r pull-prices}
prices_raw <- tq_get(
  TICKERS,
  from = TRAIN_START,
  to   = TEST_END,
  get  = "stock.prices"
)
```

## Ticker Availability

XLC (Communication Services) launched on 2018-06-19 and XLRE (Real Estate) on 2015-10-07. These tickers return fewer observations than the others. An eligibility check ensures they are excluded from portfolio allocation before their respective inception dates.

```{r ticker-availability}
#| tbl-cap: "Observation counts and first available date per ticker"

ticker_meta <- prices_raw |>
  group_by(symbol) |>
  summarise(
    first_date = min(date),
    last_date  = max(date),
    n_obs      = n(),
    .groups    = "drop"
  ) |>
  mutate(sector = SECTOR_NAMES[symbol])

ticker_meta |>
  select(symbol, sector, first_date, last_date, n_obs) |>
  gt() |>
  tab_header(title = "Ticker Availability") |>
  cols_label(
    symbol     = "Ticker",
    sector     = "Sector",
    first_date = "First Date",
    last_date  = "Last Date",
    n_obs      = "Observations"
  ) |>
  cols_align("left")
```

## Data Quality Checks

We screen for: (1) NA adjusted prices, (2) zero-volume days, and (3) single-day price moves exceeding 50% (likely data errors despite using adjusted prices).

```{r data-quality}
#| tbl-cap: "Data quality flags per ticker"

quality_flags <- prices_raw |>
  group_by(symbol) |>
  summarise(
    na_price      = sum(is.na(adjusted)),
    zero_volume   = sum(volume == 0, na.rm = TRUE),
    extreme_moves = sum(
      abs(adjusted / lag(adjusted) - 1) > 0.50,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

quality_flags |>
  gt() |>
  tab_header(title = "Data Quality Flags") |>
  cols_label(
    symbol        = "Ticker",
    na_price      = "NA Prices",
    zero_volume   = "Zero-Volume Days",
    extreme_moves = ">50% Daily Move"
  ) |>
  data_color(
    columns = c(na_price, zero_volume, extreme_moves),
    method  = "numeric",
    palette = c("white", "orange")
  )
```

## Daily Returns

```{r daily-returns}
returns_daily <- prices_raw |>
  group_by(symbol) |>
  tq_transmute(
    select     = adjusted,
    mutate_fun = periodReturn,
    period     = "daily",
    col_rename = "daily_ret"
  ) |>
  ungroup()
```

## Monthly Returns

Monthly returns are used for rebalancing signal computation and portfolio performance tracking. We use end-of-month closing prices.

```{r monthly-returns}
returns_monthly <- prices_raw |>
  group_by(symbol) |>
  tq_transmute(
    select     = adjusted,
    mutate_fun = periodReturn,
    period     = "monthly",
    col_rename = "monthly_ret"
  ) |>
  ungroup() |>
  # date is set to last trading day of each month by tq_transmute
  rename(month_end = date)
```

## Price History Visualization

```{r price-chart}
#| fig-cap: "Adjusted price history for all tickers (rebased to 100 at each series start)"

rebased <- prices_raw |>
  group_by(symbol) |>
  mutate(rebased = adjusted / first(adjusted) * 100) |>
  ungroup() |>
  mutate(sector = SECTOR_NAMES[symbol])

p <- ggplot(rebased, aes(x = date, y = rebased, color = symbol)) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  geom_vline(xintercept = as.numeric(TEST_START),
             linetype = "dashed", color = "grey40") +
  annotate("text", x = TEST_START, y = max(rebased$rebased, na.rm = TRUE) * 0.95,
           label = "Test period →", hjust = -0.1, size = 3.5, color = "grey40") +
  scale_y_continuous(labels = label_comma()) +
  scale_color_brewer(palette = "Paired") +
  labs(title = "Rebased Price History (base = 100 at inception)",
       x = NULL, y = "Rebased Price", color = "Ticker") +
  theme_minimal(base_size = 12)

ggplotly(p, tooltip = c("x", "y", "colour")) |>
  layout(hovermode = "x unified")
```

```{r returns-summary}
#| tbl-cap: "Annualized return and volatility over the full training period"

train_summary <- returns_daily |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  group_by(symbol) |>
  summarise(
    ann_return = prod(1 + daily_ret, na.rm = TRUE)^(252 / n()) - 1,
    ann_vol    = sd(daily_ret, na.rm = TRUE) * sqrt(252),
    sharpe     = ann_return / ann_vol,
    .groups    = "drop"
  ) |>
  mutate(sector = SECTOR_NAMES[symbol]) |>
  arrange(desc(ann_return))

train_summary |>
  select(symbol, sector, ann_return, ann_vol, sharpe) |>
  gt() |>
  tab_header(title = "Training Period Summary (Jun 2000 – Dec 2024)") |>
  cols_label(
    symbol     = "Ticker",
    sector     = "Sector",
    ann_return = "Ann. Return",
    ann_vol    = "Ann. Volatility",
    sharpe     = "Sharpe (rf=0)"
  ) |>
  fmt_percent(columns = c(ann_return, ann_vol), decimals = 1) |>
  fmt_number(columns = sharpe, decimals = 2) |>
  data_color(
    columns = sharpe,
    method  = "numeric",
    palette = c("tomato", "white", "steelblue")
  )
```
